---
title: "HUDK4051: Prediction - Comparing Trees"
author: "Nicole Schlosberg"
date: "2/22/21"
output: html_document
---

In this assignment you will modelling student data using three flavors of tree algorithm: CART, C4.5 and C5.0. We will be using these algorithms to attempt to predict which students drop out of courses. Many universities have a problem with students over-enrolling in courses at the beginning of semester and then dropping most of them as the make decisions about which classes to attend. This makes it difficult to plan for the semester and allocate resources. However, schools don't want to restrict the choice of their students. One solution is to create predictions of which students are likely to drop out of which courses and use these predictions to inform semester planning. 

In this assignment we will be using the tree algorithms to build models of which students are likely to drop out of which classes. 

## Software

In order to generate our models we will need several packages. The first package you should install is [caret](https://cran.r-project.org/web/packages/caret/index.html).

There are many prediction packages available and they all have slightly different syntax. caret is a package that brings all the different algorithms under one hood using the same syntax. 

We will also be accessing an algorithm from the [Weka suite](https://www.cs.waikato.ac.nz/~ml/weka/). Weka is a collection of machine learning algorithms that have been implemented in Java and made freely available by the University of Waikato in New Zealand. To access these algorithms you will need to first install both the [Java Runtime Environment (JRE) and Java Development Kit](http://www.oracle.com/technetwork/java/javase/downloads/jre9-downloads-3848532.html) on your machine. You can then then install the [RWeka](https://cran.r-project.org/web/packages/RWeka/index.html) package within R.

**Weka requires Java and Java causes problems. If you cannot install Java and make Weka work, please follow the alternative instructions at line 121**
(Issue 1: failure to install RWeka/RWekajars, paste "sudo R CMD javareconf" into terminal and try to install again)

The last package you will need is [C50](https://cran.r-project.org/web/packages/C50/index.html).

## Data

The data comes from a university registrar's office. The code book for the variables are available in the file code-book.txt. Examine the variables and their definitions.

Upload the drop-out.csv data into R as a data frame. 

```{r}
DATA <- read.csv("drop-out.csv", header = TRUE)
```

The next step is to separate your data set into a training set and a test set. Randomly select 25% of the students to be the test data set and leave the remaining 75% for your training data set. (Hint: each row represents an answer, not a single student.)

```{r}
library(caret)

trainData <- createDataPartition( y = DATA$student_id, p = .75, list = FALSE)
TRAIN1 <- DATA[trainData,]
TEST1 <- DATA[-trainData,]
```

For this assignment you will be predicting the student level variable "complete". 
(Hint: make sure you understand the increments of each of your chosen variables, this will impact your tree construction)

Visualize the relationships between your chosen variables as a scatterplot matrix.  Save your image as a .pdf named scatterplot_matrix.pdf. 

```{r}
library(tidyr)
library(dplyr)
DATApdf <- DATA %>% mutate(complete=ifelse(complete=="yes", 1,0)) %>% mutate(international=ifelse(international=="yes", 1,0)) %>% mutate(online=ifelse(online=="yes", 1,0))

pdf("scatterplot_matrix.pdf")
pairs(DATApdf)
dev.off()
```
Based on this visualization do you see any patterns of interest? Why or why not?

ANSWER: One pattern is between years and complete. Students who have spent less time in the program have completed more courses than those who have spent more amount time in their program; years and complete are negatively related.

## CART Trees

You will use the [rpart package](https://cran.r-project.org/web/packages/rpart/rpart.pdf) to generate CART tree models.

Construct a classification tree that predicts complete using the caret package.

```{r}
library(caret)
library(rpart)

TRAIN2 <- TRAIN1[,c(2:10)] #Remove the student_id variable that we do not want to use in the model

#caret does not summarize the metrics we want by default so we have to modify the output
MySummary  <- function(data, lev = NULL, model = NULL){
  df <- defaultSummary(data, lev, model)
  tc <- twoClassSummary(data, lev, model)
  pr <- prSummary(data, lev, model)
  out <- c(df,tc,pr)
  out}

#Define the control elements we would like to use
ctrl <- trainControl(method = "repeatedcv", #Tell caret to perform k-fold cross validation
                repeats = 3, #Tell caret to repeat each fold three times
                classProbs = TRUE, #Calculate class probabilities
                summaryFunction = MySummary)

#Define the model
cartFit <- train(complete ~ ., #Define which variable to predict 
                data = TRAIN2, #Define the data set to train the model on
                trControl = ctrl, #Tell caret the control elements
                method = "rpart", #Define the model type
                metric = "Accuracy", #Final model choice is made according to sensitivity
                preProc = c("center", "scale")) #Center and scale the data to minimize the
               
#Check the results
cartFit
```

Describe important model attributes of your tree. Do you believe it is a successful model of student performance, why/why not?

ANSWER: The final value of cp that was used for this model was 0.0123. The corresponding value of Accuracy to this cp is 0.8923, which suggests that there is a 89.23% probability of a randomly selected student from a "completed" group being classified as "completed" as opposed to a randomly selected student from a "not completed" group being classified as "completed", which is fairly high. 

Can you use the sensitivity and specificity metrics to calculate the F1 metric?

ANSWER: The specificity (true negative rate) of the model is 0.9955, suggesting that the rate of a correct classification of students who have not completed a course is 99.55%. However, the sensitivity (true positive rate) of the model is 0.6577, indicating that the rate of a correct classification of students who have completed a course is only 65.77%. Some of those who have completed the course, where incorrectly classified as "not completed". F1 = 2TP/(2TP+FP+FN), so yes you can figure out the F1 measure because you know the TP and TN (so add them together and then subtract from the total and that equals FP+FN). When you use the formula you can see that the models performance could be more successful.

Now predict results from the test data and describe important attributes of this test. Do you believe it is a successful model of student performance, why/why not?

```{r}
TEST2 <- TEST1[,c(2:10)] #Remove the student_id variable that we do not want to use in the model

#Generate prediction using previously trained model
cartClasses <- predict(cartFit, newdata = TEST2)

#Generate model statistics
confusionMatrix(data = cartClasses, as.factor(TEST2$complete))
```
ANSWER: The Accuracy of the model is 0.905 (90.50%). Specificity is 0.9971 (99.71%). Thus the rate of a successful prediction of students not having completed a course is high. However, the value of sensitivity is 0.6754 (67.54%), meaning the rate of a successful prediction of students who have completed a course is only 67.54%, so the model can be improved.


## Conditional Inference Trees

Train a Conditional Inference Tree using the `party` package on the same training data and examine your results.
```{r}
library(party)
condFit <- train(complete ~ .,
                data = TRAIN2,
                trControl = ctrl,
                method = "ctree",
                metric = "Accuracy",
                preProc = c("center", "scale"))
condFit
plot(condFit)
```
Describe important model attributes of your tree. Do you believe it is a successful model of student performance, why/why not?

ANSWER: The accuracy only represents a global measure. It is a good first look, but it hides TP and TN rate. 

What does the plot represent? What information does this plot tell us?

ANSWER: The plot represents that Accuracy against 1 - P-Value Threshold. It spans the 3  repetitions. The plot tells us for each percentage of accuracy (88.35%, 89.41%,89.52%) what mincriterion, the value of the test statistic or 1 - p-value that must be exceeded in order to implement a split in the tree, corresponds (1%,50%,99%). In other words the p-value where the tree splits.

Now test your new Conditional Inference model by predicting the test data and generating model fit statistics.

```{r}
condFitClasses <- predict(condFit, newdata = TEST2)
confusionMatrix(data = condFitClasses, as.factor(TEST2$complete))
```

There is an updated version of the C4.5 model called C5.0, it is implemented in the C50 package. 
What improvements have been made to the newer version? 

ANSWER: C5.0 algorithm automatically winnows the data before creating classifiers; it purposefully removes the predictors that can improve the accuracy of the model. It has boosting iterations technique. Also, it provides rules and trees for its model types. The rule-based model breaks down the constructed tree into mutually exclusive rules rather than the tree-based model.

Install the C50 package, train and then test the C5.0 model on the same data.
```{r}
library(C50)
c50Fit <-  train(complete ~ .,
                data = TRAIN2,
                trControl = ctrl,
                method = "C5.0",
                metric = "Accuracy",
                preProc = c("center", "scale"))
c50Fit
plot(c50Fit)
```


## Compare the models

caret allows us to compare all three models at once.

```{r}
resamps <- resamples(list(cart = cartFit, condinf = condFit, cfiveo = c50Fit))
summary(resamps)
```
What does the model summary tell us? Which model do you believe is the best?

ANSWER: The model summary shows the summary descriptive statistics of many metrics for each one of the three models. Based on the model summary the following models have the highest average metrics: Accuracy- Conditional Inference Tree; AUC- C50; F- cart; Kappa- Conditional Inference Tree; Precision- cart; Recall- C50; ROC- C50; Sensitivity- C50; Specificity- cart. When only looking at specificity and sensitivity the C50 model has the best measures meaning that it predicts the drop out better than the other trees. However the average Sensitivity is still low (67.27%) which means the model could still do better. 

Which variables (features) within your chosen model are important, do these features provide insights that may be useful in solving the problem of students dropping out of courses?

ANSWER: Years (the number of years enrolled in a program) from the drop-out.csv data greatly predicts whether the student will complete the course or dropout. As seen in the scatterplot complete was related to the number of years in their program. 